Multi-Core Model Checking and Maximum Satisfiability Applied to Hardware-Software Partitioning

### 2016-08-15 Docs/IJES/journal.tex
revised version of paper

Starting at the $90$'s decade, intensive research was performed in HW-SW partitioning, and several approaches proposed, as shown in~\cite{Arato2003} and~\cite{Mann2007}. In~\cite{Trindade2015} and~\cite{Trindade2016} was shown that it is possible to use Bounded Model Checking (BMC) based on Satisfiability Modulo Theories (SMT), implemented in a tool called Efficient SMT-Based Context-Bounded Model Checker (ESBMC) as defined in~\cite{Cordeiro2012}, in order to perform HW-SW partitioning in embedded systems. In~\cite{Trindade2016}, the original ESBMC algorithm was improved in terms of performance, so a multi-core sequential approach was introduced, in order to speed up the performance and turn it similar, in some cases, to conventional optimization tools, which are used to perform HW-SW partitioning.

The present work thus extends those previous studies since there is a substantial improvement in terms of the SMT-based verification methods. In particular, we extend the work of~\cite{Trindade2015} and~\cite{Trindade2016} to exploit a multi-core parallel- and binary-search approaches, as well as, the integration of the Maximum SMT (MaxSMT) solver $\nu$Z into ESBMC, which is a state-of-art optimization tool based on SMT as defined in~\cite{Bjorner2015}.

We exploit the availability of multi-core processors and the tool OpenMP as defined in~\cite{Tang2009}; thus, we apply SMT-based verification methods to the HW-SW partition problem in three different ways using a multi-core ESBMC approach with OpenMP: ESBMC-SS using a sequential-search (SS), ESBMC-PS using a parallel-search (PS), and ESBMC-PB using a binary-search (BS). Experimental results are compared to ILP (integer linear programming), GA (generic algorithms) in a multi-core version, and also to $\nu$Z, which supports only a single-core approach, according to~\cite{Bjorner2015}. The ILP and GA algorithms are implemented with the optimization toolbox of Matlab, according to~\cite{TheMathWorks2013}, while $\nu$Z is a built-in tool to the SMT solver Z3. All multi-core ESBMC approaches, together with $\nu$Z, are implemented with the ESBMC tool.

\subsection{Contributions}

The main contribution of the present study is to describe and evaluate a comprehensive SMT-based BMC approach in a multi-core architecture applied to solve HW-SW optimization problems. Additionally, we integrate the MaxSMT solver $\nu$Z into an off-the-shelf BMC tool, which is typically used for software verification, in order to formulate and solve optimization problems within the logical context of constraints. Experimental results show that multi-core model-checking techniques can be effective, in particular cases, to find the optimal solution of the HW-SW partitioning problem using an SMT-based BMC approach. Although there is no single tool for efficiently solving all HW-SW partitioning problems, we show that the MaxSMT solver $\nu$Z is faster than other state-of-the-art optimization tools for small- and medium-size optimization problems.  To the best of our knowledge, this is the first work to use a multi-core SMT-based verification and a MaxSMT solver to check for HW-SW partitioning problems in embedded systems.

@@

Remembering that, as defined by formal model, there is a index of each decision variable, indicated by letter "i", which ranges from zero to the number of nodes of the problem to be solved.

@@

%----------------------------------------------
\section{Related Work}
\label{Related-Work}
%----------------------------------------------

Since the second half of the first decade of 2000s, three main paths have been tracked to improve or to present alternative solutions to the optimization of HW-SW partitioning, {\it i.e.}, to find the exact solution, as shown in~\cite{Mann2007}; to use heuristics to speed up performance time, as shown in~\cite{Arato2003}; and hybrid ones, as shown in~\cite{Arato2005}.

In the first group, the exact solution to the HW-SW partitioning problem is found. The use of SMT-based verification presented in this paper can be grouped into this category, because the exact solution is found with the given algorithms. The difference is based only on the technique chosen to solve the problem.
Another path followed in past initiatives and which has had more studies is the creation of heuristics to speed up the running time of the solution. The difference of this kind of solution to SMT-based verification and maximum satisfiability is based on two facts: all ESBMC implementations guarantee to find the exact solution, but heuristics are faster, when the complexity is greater.

Finally, there are approaches that mixes heuristics with exact solution tools. The idea is to use a heuristic to speed up some phase of an exact solution tool. \textcolor{blue}{It is worth mentioning} that the final solution is not necessarily an optimal global solution. Only the SMT-based verification is guaranteed to find the exact solution, but hybrid algorithms are faster when complexity rises.

In terms of SMT-based verification, most related studies are restricted to present the model, its modification to programming languages ({\it e.g.}, C/C++ and Java), and the application to multi-thread algorithms or to embedded systems to check for program correctness. In~\cite{Ramalho2013} it presents a bounded model checker for C++ programs, which is an evolution of dealing with C programs, and~\cite{Cordeiro2012} uses ESBMC for embedded ANSI-C software. In~\cite{Trindade2015}, and~\cite{Trindade2016} it was proven that it is possible to use ESBMC to solve HW-SW partitioning in a single- and multi-core way, but the former has performance issues that were improved by the latter, which used only a sequential search to perform multi-core model checking. There are related studies focused on decreasing the verification time of model checkers by applying Swarm Verification, as shown in~\cite{Holzmann2011}, and modifications of internal search engines to add support for parallelism, as shown in~\cite{Holzmann2012}, but there is still the need for initiatives related to parallel SMT solver, according to~\cite{Wintersteiger2009}.

Recently, the SMT solver Z3 has been extended to pose and solve optimization problems modulo theories, as shown in~\cite{Bjorner2015}. In particular, $\nu$Z tool offers substantial performance improvement in optimization problems, according to~\cite{Bjorner2014} and~\cite{Bjorner2015}. As an application example, \cite{Pavlinovic2015} propose an approach which considers all possible compiler error sources for statically typed functional programming languages and reports the most useful one subject to some usefulness criterion. The authors formulate this approach as an optimization problem related to SMT and use $\nu$Z to compute an optimal error source in a given ill-typed program. The approach described by \cite{Pavlinovic2015}, which uses MaxSMT solver $\nu$Z, shows a significant performance improvement if compared to previous SMT encodings and localization algorithms.

The problem addressed in this present paper uses a single objective function for minimization. In~\cite{Patrick2015}, OptiMathSAT and $\nu$Z are compared for software optimization problems. OptiMathSAT, using multiple objective functions, works better than $\nu$Z; however, for problems with a single objective function, the performance of $\nu$Z is better than OptiMathSAT.

>>>>

%-----------------related work came to here and was rewriten
\textcolor{blue}{Since the first decade of 2000s, two main paths have been tracked to solve the HW-SW partitioning problem, {\it i.e.}, to find the exact solution of a optimization problem, as shown in~\cite{Mann2007}; to use heuristics to speed up performance time, as shown in~\cite{Arato2003} and~\cite{Arato2005}. It is worth mentioning that the using heuristics the final solution is not necessarily an optimal global solution.}

\textcolor{blue}{In terms of SMT-based verification, most related studies are restricted to present the model, its modification to programming languages ({\it e.g.}, C/C++ and Java), and the application to multi-thread algorithms or to embedded systems to check for program correctness. In~\cite{Ramalho2013} it presents a bounded model checker for C++ programs, which is an evolution of dealing with C programs, and~\cite{Cordeiro2012} uses ESBMC for embedded ANSI-C software. In~\cite{Trindade2015}, and~\cite{Trindade2016} it was proven that it is possible to use ESBMC to solve HW-SW partitioning in a single- and multi-core way, but the former has performance issues that were improved by the latter, which used only a sequential search to perform multi-core model checking. There are related studies focused on decreasing the verification time of model checkers by applying Swarm Verification, as shown in~\cite{Holzmann2011}, and modifications of internal search engines to add support for parallelism, as shown in~\cite{Holzmann2012}, but there is still the need for initiatives related to parallel SMT solver, according to~\cite{Wintersteiger2009}.}

\textcolor{blue}{Recently, the SMT solver Z3 has been extended to pose and solve optimization problems modulo theories, as shown in~\cite{Bjorner2015}. In particular, $\nu$Z tool offers substantial performance improvement in optimization problems, according to~\cite{Bjorner2014} and~\cite{Bjorner2015}. As an application example, \cite{Pavlinovic2015} propose an approach which considers all possible compiler error sources for statically typed functional programming languages and reports the most useful one subject to some usefulness criterion. The authors formulate this alternative single-core approach as an optimization problem related to SMT and use $\nu$Z to compute an optimal error source in a given ill-typed program.}

\textcolor{blue}{At this work, we apply SMT-based verification methods to the HW-SW partition problem in three different ways using a multi-core ESBMC approach with OpenMP: ESBMC-SS using a sequential-search (SS), ESBMC-PS using a parallel-search (PS), and ESBMC-PB using a binary-search (BS). Experimental results are compared to ILP (integer linear programming), GA (generic algorithms) in a multi-core version, and also to $\nu$Z, which supports only a single-core approach, according to~\cite{Bjorner2015}. The ILP and GA algorithms are implemented with the optimization toolbox of Matlab, according to~\cite{TheMathWorks2013}, while $\nu$Z is a built-in tool to the SMT solver Z3. All multi-core ESBMC approaches, together with $\nu$Z, are implemented with the ESBMC tool. To the best of our knowledge, this is the first work to use a multi-core SMT-based verification and a MaxSMT solver to check for HW-SW partitioning problems in embedded systems.}

%-----------------
%Starting at the $90$'s decade, intensive research was performed in HW-SW partitioning, and several approaches proposed, as shown in~\cite{Arato2003} and~\cite{Mann2007}. In~\cite{Trindade2015} and~\cite{Trindade2016} was shown that it is possible to use Bounded Model Checking (BMC) based on Satisfiability Modulo Theories (SMT), implemented in a tool called Efficient SMT-Based Context-Bounded Model Checker (ESBMC) as defined in~\cite{Cordeiro2012}, in order to perform HW-SW partitioning in embedded systems. In~\cite{Trindade2016}, the original ESBMC algorithm was improved in terms of performance, so a multi-core sequential approach was introduced, in order to speed up the performance and turn it similar, in some cases, to conventional optimization tools, which are used to perform HW-SW partitioning.

%The present work thus extends those previous studies since there is a substantial improvement in terms of the SMT-based verification methods. In particular, we extend the work of~\cite{Trindade2015} and~\cite{Trindade2016} to exploit a multi-core parallel- and binary-search approaches, as well as, the integration of the Maximum SMT (MaxSMT) solver $\nu$Z into ESBMC, which is a state-of-art optimization \textcolor{blue}{single-core} tool based on SMT as defined in~\cite{Bjorner2015}.

%We exploit the availability of multi-core processors and the tool OpenMP as defined in~\cite{Tang2009}; thus, we apply SMT-based verification methods to the HW-SW partition problem in three different ways using a multi-core ESBMC approach with OpenMP: ESBMC-SS using a sequential-search (SS), ESBMC-PS using a parallel-search (PS), and ESBMC-PB using a binary-search (BS). Experimental results are compared to ILP (integer linear programming), GA (generic algorithms) in a multi-core version, and also to $\nu$Z, which supports only a single-core approach, according to~\cite{Bjorner2015}. The ILP and GA algorithms are implemented with the optimization toolbox of Matlab, according to~\cite{TheMathWorks2013}, while $\nu$Z is a built-in tool to the SMT solver Z3. All multi-core ESBMC approaches, together with $\nu$Z, are implemented with the ESBMC tool.

%The main contribution of the present study is to describe and evaluate a comprehensive SMT-based BMC approach in a multi-core architecture applied to solve HW-SW optimization problems. Additionally, we integrate the MaxSMT solver $\nu$Z into an off-the-shelf BMC tool, which is typically used for software verification, in order to formulate and solve optimization problems within the logical context of constraints. Experimental results show that multi-core model-checking techniques can be effective, in particular cases, to find the optimal solution of the HW-SW partitioning problem using an SMT-based BMC approach. Although there is no single tool for efficiently solving all HW-SW partitioning problems, we show that the MaxSMT solver $\nu$Z is faster than other state-of-the-art optimization tools for small- and medium-size optimization problems.  To the best of our knowledge, this is the first work to use a multi-core SMT-based verification and a MaxSMT solver to check for HW-SW partitioning problems in embedded systems.

%\subsection{Contributions}

%The main contribution of the present study is to describe and evaluate a comprehensive SMT-based BMC approach in a multi-core architecture applied to solve HW-SW optimization problems. Additionally, we integrate the MaxSMT solver $\nu$Z into an off-the-shelf BMC tool, which is typically used for software verification, in order to formulate and solve optimization problems within the logical context of constraints. Experimental results show that multi-core model-checking techniques can be effective, in particular cases, to find the optimal solution of the HW-SW partitioning problem using an SMT-based BMC approach. Although there is no single tool for efficiently solving all HW-SW partitioning problems, we show that the MaxSMT solver $\nu$Z is faster than other state-of-the-art optimization tools for small- and medium-size optimization problems.  To the best of our knowledge, this is the first work to use a multi-core SMT-based verification and a MaxSMT solver to check for HW-SW partitioning problems in embedded systems.

@@

\textcolor{blue}{Remembering that, as defined by formal model, there is a index of each decision variable, indicated by letter "i", which ranges from zero to the number of nodes of the problem to be solved.}

@@

%----------------------------------------------
%\section{Related Work}
%\label{Related-Work}
%----------------------------------------------

%Since the second half of the first decade of 2000s, three main paths have been tracked to improve or to present alternative solutions to the optimization of HW-SW partitioning, {\it i.e.}, to find the exact solution, as shown in~\cite{Mann2007}; to use heuristics to speed up performance time, as shown in~\cite{Arato2003}; and hybrid ones, as shown in~\cite{Arato2005}.

%In the first group, the exact solution to the HW-SW partitioning problem is found. The use of SMT-based verification presented in this paper can be grouped into this category, because the exact solution is found with the given algorithms. The difference is based only on the technique chosen to solve the problem.
%Another path followed in past initiatives and which has had more studies is the creation of heuristics to speed up the running time of the solution. The difference of this kind of solution to SMT-based verification and maximum satisfiability is based on two facts: all ESBMC implementations guarantee to find the exact solution, but heuristics are faster, when the complexity is greater.

%Finally, there are approaches that mixes heuristics with exact solution tools. The idea is to use a heuristic to speed up some phase of an exact solution tool. \textcolor{blue}{It is worth mentioning} that the final solution is not necessarily an optimal global solution. Only the SMT-based verification is guaranteed to find the exact solution, but hybrid algorithms are faster when complexity rises.

%In terms of SMT-based verification, most related studies are restricted to present the model, its modification to programming languages ({\it e.g.}, C/C++ and Java), and the application to multi-thread algorithms or to embedded systems to check for program correctness. In~\cite{Ramalho2013} it presents a bounded model checker for C++ programs, which is an evolution of dealing with C programs, and~\cite{Cordeiro2012} uses ESBMC for embedded ANSI-C software. In~\cite{Trindade2015}, and~\cite{Trindade2016} it was proven that it is possible to use ESBMC to solve HW-SW partitioning in a single- and multi-core way, but the former has performance issues that were improved by the latter, which used only a sequential search to perform multi-core model checking. There are related studies focused on decreasing the verification time of model checkers by applying Swarm Verification, as shown in~\cite{Holzmann2011}, and modifications of internal search engines to add support for parallelism, as shown in~\cite{Holzmann2012}, but there is still the need for initiatives related to parallel SMT solver, according to~\cite{Wintersteiger2009}.

%Recently, the SMT solver Z3 has been extended to pose and solve optimization problems modulo theories, as shown in~\cite{Bjorner2015}. In particular, $\nu$Z tool offers substantial performance improvement in optimization problems, according to~\cite{Bjorner2014} and~\cite{Bjorner2015}. As an application example, \cite{Pavlinovic2015} propose an approach which considers all possible compiler error sources for statically typed functional programming languages and reports the most useful one subject to some usefulness criterion. The authors formulate this approach as an optimization problem related to SMT and use $\nu$Z to compute an optimal error source in a given ill-typed program. The approach described by \cite{Pavlinovic2015}, which uses MaxSMT solver $\nu$Z, shows a significant performance improvement if compared to previous SMT encodings and localization algorithms.

%The problem addressed in this present paper uses a single objective function for minimization. In~\cite{Patrick2015}, OptiMathSAT and $\nu$Z are compared for software optimization problems. OptiMathSAT, using multiple objective functions, works better than $\nu$Z; however, for problems with a single objective function, the performance of $\nu$Z is better than OptiMathSAT.
